---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gasExchangeR

<!-- badges: start -->
<!-- badges: end -->

The goal of gasExchangeR is to help exercise physiologist more easily analyze gas exchange data and to provide more directly control when doing so. The overview of the process of analyzing gas exchange data involves importing the data, identifying and removing outliers, interpolating if desired, and finally averaging the data. Afterwards then one can determine ventilatory thresholds, VO2max, and other important values.

## Installation

You can (hopefully) install this package from CRAN. If that doesn't work, you can install the development version of gasExchangeR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ahesse2567/gasExchangeR")
```

## Development notes

The gasExchangeR package is part of [Anton Hesse's](https://orcid.org/0000-0001-8456-7343) PhD dissertation at the University of Minnesota-Twin Cities. Please expect changes and improvements if you use this package, and please submit feedback if you encounter bugs or have suggestions.

## Early acknowledgemets and other recommended packages

This package expands on the work by Felipe Mattioni Maturana (https://orcid.org/0000-0002-4221-6104). Specifically, his work on the [whippr](https://github.com/fmmattioni/whippr) and [lacater](https://fmmattioni.github.io/lactater/) packages. [gasExchangeR](https://github.com/ahesse2567/gasExchangeR) focuses more on graded exercise testing than VO2 kinetics; it also emphasizes ventilatory breakpoint algorithms.

## Using this package to find ventilatory thresholds

Many exercise studies require finding the first and second ventilatory thresholds (VT1 & VT2). However, breath-by-breath data is highly variable and requires some cleaning prior to finding these thresholds.

```{r example, message=FALSE}
# Load libraries
library(gasExchangeR)
library(tidyverse)
library(janitor)
```

### Data Processing

```{r initial tidying, warning=FALSE}
# read in raw data
file_lines <- readLines("inst/extdata/Anton_vo2max.txt")
df_raw <- read.table(textConnection(file_lines[-2]), header = TRUE, sep="\t")

# initial data tidying
df_unavg <- df_raw %>%
    as_tibble() %>%
    clean_names() %>%
    separate(`time`, into = c("m1", "s1"), sep = ":") %>%
    separate(ex_time, into = c("m2", "s2"), sep = ":") %>%
    separate(time_clock,
             into = c("h3", "m3", "s3"),
             sep = ":") %>%
    mutate(across(where(is.character), as.numeric)) %>%
    mutate(time = (m1*60 + s1), .keep = "unused") %>%
    mutate(ex_time = (m2*60 + s2 ), .keep = "unused") %>%
    mutate(clock_time = hms::hms(s3, m3, h3), .keep = "unused") %>%
    relocate(contains("time")) %>%
    filter(!is.na(ex_time)) %>%
    filter(speed >= 4.5 & ex_time >= 750) %>%
    select(-time) %>%
    rename(time = ex_time,
           vo2_kg = vo2,
           vo2 = vo2_1,
           ve = ve_btps) %>%
    # calculate common variables
    mutate(ve_vo2 = ve / vo2 * 1000, 
           ve_vco2 = ve/vco2*1000,
           excess_co2 = vco2^2 / vo2 - vco2)
```

Plotting the raw data

```{r plot_raw_data}
ggplot(data = df_unavg, aes(x = time)) +
  geom_point(aes(y = vo2, color = "vo2"), alpha = 0.5) +
  geom_point(aes(y = vco2, color = "vco2"), alpha = 0.5) +
  scale_color_manual(values = c("vo2" = "red", "vco2" = "blue")) +
  theme_minimal() +
  ggtitle("Breath-by-breath (Raw) Data") +
  scale_color_discrete(name = "", labels = c("vo2" = "vo2", "vco2" = "vco2"))

```

The raw data is obviously noisy. We will first use a rolling-breath average with absolute VO2 values to remove outliers.

```{r warning=FALSE}
df_unavg_no_outliers <- df_unavg %>% 
    ventilatory_outliers(outlier_cols = "vo2", max_passes = 1,
                         plot_outliers = TRUE)
```

Removing outliers helps, but some averaging is also required.

```{r}
df_avg <- df_unavg_no_outliers %>% 
    avg_exercise_test(method = "time", calc_type = "bin", bin_w = 10)

ggplot(data = df_avg, aes(x = time)) +
  geom_point(aes(y = vo2, color = "vo2"), alpha = 0.5) +
  geom_point(aes(y = vco2, color = "vco2"), alpha = 0.5) +
  scale_color_manual(values = c("vo2" = "red", "vco2" = "blue")) +
  theme_minimal() +
  ggtitle("10-Second Bin-Average") +
  scale_color_discrete(name = "", labels = c("vo2" = "vo2", "vco2" = "vco2"))

```
### Finding Ventilatory Thresholds

```{r warning=FALSE}
bp_dat <- breakpoint(.data = df_avg, method = "v-slope",
                     algorithm_vt2 = "d2_reg_spline_maxima",
                     x_vt2 = "vo2", y_vt2 = "ve_vco2",
                     vo2 = "vo2", vco2 = "vco2", ve = "ve", time = "time",
                     bp = "both", truncate = TRUE, 
                     front_trim_vt1 = 60, front_trim_vt2 = 60,
                     pos_change_vt1 = TRUE, pos_change_vt2 = TRUE,
                     pos_slope_after_bp = TRUE)
print(bp_dat$bp_dat, width = Inf)
bp_dat$vt1_dat$bp_plot
bp_dat$vt2_dat$bp_plot
```


